---
title: "EPI 207 Assignment 2"
author: ""
output: html_document
date: "10 Feb 2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
getwd()
library(haven)
library(tidyverse)
library(dplyr)
library(labelled)
```

#Loading CHIS data
```{r}
adult_chis_2023 <- read_sas("C:/Users/sonal/Desktop/2023 Adult SAS/adult.sas7bdat") 
```

#Cleaning CHIS data
```{r}
#Select variables from original data
clean_ogname_data <- adult_chis_2023 %>% 
  #select variables of interest
  select(SRAGE_P1, #age
         SRSEX,   #gender
         OMBSRR_P1, #race
         BINGE30, #binge drinking
         DISTRESS,#SPD
         DSTRS_P1, #SPD (PUF 1 year recode)
         DSTRS30, #likely SPD in past month
         DSTRS12, #likely SPD in past year
         AHEDC_P1,  #education 
         WRKST_P1,   #work status
         POVLL,   #poverty level
         POVLL2_P1V2, #poverty level - numeric
         SRTENR, #self-reported household tenure
         AH44A, #used paid childcare in past month
         MARIT,   #marital status - 3 cat
         MARIT2,  #marital status - 4 cat
         MARIT_45, #marital status - 45+
         RAKEDW0, #weights
         PROXY, #proxy interview
         AF81, #needed help for emot/mntl or alc/drug problem in past year
         FAMT4 #marital/parental status
         )
```

```{r}
#rename variables and create new ones
renamed_data <- clean_ogname_data %>%
#rename variables with more intuitive names
  mutate(age           = SRAGE_P1, #age
         gender        = SRSEX,   #gender
         race          = OMBSRR_P1, #race
         binge         = BINGE30, #Binge Drinking
         spd           = DISTRESS,#SPD
         spd_month     = DSTRS30, #SPD in past month
         spd_year      = DSTRS12, #SPD in past year
         edu           = AHEDC_P1,  #education 
         emply_status  = WRKST_P1,   #work status
         poverty       = POVLL,  #poverty
         poverty_fpl   = POVLL2_P1V2, #poverty level - numeric
         tenure        = SRTENR, #self-reported household tenure
         childcare     = AH44A, #used paid childcare in past month
         marital       = MARIT,   #marital status - 3 cat
         marital2      = MARIT2,  #marital status - 4 cat
         marital45     = MARIT_45, #marital status - 45+
         weight        = RAKEDW0, #weights
         proxy         = PROXY, #proxy interview
         subuse_mental = AF81, #needed help for emot/mntl or alc/drug problem in past year
         fam_type      = FAMT4,
    
         #create/edit variables to analyze
         age_cat = case_when(
          age >= 18 & age < 26 ~ "18–25 (reference)",
          age >= 26 & age < 35 ~ "26–34",
          age >= 35 & age < 45 ~ "35–44",
          age >= 45 & age < 55 ~ "45-54",
          age >= 55 & age < 65 ~ "55-64",
          age >= 65 ~ "65+",
          TRUE ~ as.character(NA)),
         
         gender = case_when(
           gender == 1 ~ "Male",
           gender == 2 ~ "Female",
           TRUE ~ as.character(NA)),
         
         race = case_when(
           race == 1 ~ "Hispanic",
           race == 2 ~ "White, Non-Hispanic",
           race == 3 ~ "African American, Non-Hispanic",
           race == 4 ~ "American Indian/Alaskan Native, NH",
           race == 5 ~ "Asian, Non-Hispanic",
           race == 6 ~ "Other/Two or more races",
           TRUE ~ as.character(NA)),
         
          binge = case_when(
           binge == 1 ~ "Yes",
           binge == 2 ~ "No",
           TRUE ~ as.character(NA)),
         
         spd_lvl = case_when(
           spd_month == 1 ~ "Acute SPD",
           spd_month == 2 & spd_year == 1 ~ "Recent SPD",
           spd_year == 2 ~ "No SPD",
           TRUE ~ as.character(NA)),
         
         spd_month = case_when(
           spd_month == 1 ~ "Yes",
           spd_month ==2 ~ "No",
           TRUE ~ as.character(NA)),
         
         spd_year = case_when(
           spd_year == 1 ~ "Yes",
           spd_year == 2 ~ "No",
           TRUE ~ as.character(NA)),
         
         edu_lvl = case_when(
           edu == 1 ~ "<High-school degree",
           edu == 2 ~ "<High-school degree",
           edu == 3 ~ "High-school graduate (reference)",
           edu == 4 ~ "Some college",
           edu == 5 ~ "College or more",
           edu == 6 ~ "College or more",
           edu == 7 ~ "College or more",
           edu == 8 ~ "College or more",
           edu == 9 ~ "College or more"),
         
          poverty_lvl = case_when(
           poverty_fpl < 1 ~ "<100% FPL (reference)",
           poverty_fpl >= 1 & poverty_fpl < 2 ~ "100%-199% FPL",
           poverty_fpl >= 2 & poverty_fpl < 4 ~ "200%-399% FPL",
           poverty_fpl >= 4 ~ "≥ 400% FPL",
           TRUE ~ as.character(NA)),

         marital_3cat =  case_when(
          marital == 1 ~ "Married",
          marital == 2 ~ "Other",
          marital == 3 ~ "Never married",
          TRUE ~ as.character(NA)),
         
        marital_4cat =  case_when(
          marital2 == 1 ~ "Married",
          marital2 == 2 ~ "Living with partner",
          marital2 == 3 ~ "Wid/sep/div",
          marital2 == 4 ~ "Never married",
          TRUE ~ as.character(NA)), 
        
        marital_2cat =  case_when(
          marital == 1 ~ "Married",
          marital == 2 ~ "Not married",
          marital == 3 ~ "Not married",
          TRUE ~ as.character(NA)), 
        
        subuse_mental = case_when(
          subuse_mental == 1 ~ "Yes",
          subuse_mental == 2 ~ "No",
          TRUE ~ as.character (NA)),
        
        kids = case_when(
        fam_type == 1 ~ "No",
        fam_type == 2 ~ "No",
        fam_type == 3 ~ "Yes",
        fam_type == 4 ~ "Yes",
        TRUE ~ as.character (NA))
        )
```

```{r}
  
clean_data <- renamed_data %>%
  filter(!is.na(spd_lvl), !is.na(PROXY), !is.na(subuse_mental)) %>%
  #select variables needed for analysis
  #uncomment additional variables if desired
  select(age_cat,
         gender,   
         race, 
         spd_lvl,
         spd_month,
         spd_year,
         edu_lvl,
         poverty_lvl,
         marital_3cat,
         marital_2cat,
         subuse_mental,
         kids,          #has kids
         binge,         #binge drinking
         weight,
         #childcare,    #used paid childcare in past month
         #marital_4cat,   
         #marital45,
         #tenure,       #self-reported household tenure
         #proxy
         ) %>%
  #variable labeling #fix(?)
  set_variable_labels(
    age_cat = "Age (5 cat)",
    race = "Race/Ethnicity (6 cat)",
    poverty_lvl = "Poverty level (4 cat)", 
    spd_lvl = "Serious Psychological Distress (3 cat)",
    edu_lvl = "Educational Attainment (3 cat)",
    subuse_mental = "Needed Help for Substance/Mental Problem"
  )
```

#Additional Packages
```{r}
#install.packages("ggtext")
library(ggplot2)
library(ggtext)
#install.packages("table1")
library(table1)
library(tableone)
library(gt)
library(dplyr)
library(gtsummary)
library(flextable)

library(labelled)
library(codebook)
library(tibble)
#install.packages("writexl")
library(writexl)
library(here)
library(magrittr)
```

#Data Dictionary
```{r}
#Data dictionary creation 
datadic_var <- function(.data, var) {
  list_var <- .data %>%
    look_for(var) %>%
    as_tibble() %>%
    select(variable) %>%
    as.matrix() %>%
    as.vector()
  binary_var <- .data %>%
    dplyr::select(where(is_binary_var)) %>%
    colnames()
  cb <- .data %>%
    select(all_of(list_var)) %>%
    codebook::codebook_table() %>%
    dplyr::rename(`Variable name` = name,
                  Label = label,
                  `Variable type` = data_type,
                  #Values = value_labels,
                  `Complete rate` = complete_rate) %>%
    select(-c(n_missing, min, sd, median, 
              max, mean, hist, contains("format"))) %>%
    mutate(`Variable type` = 
             case_when(`Variable type`=="haven_labelled" ~ "categorical",
                       `Variable type`=="factor" ~ "categorical",
                       `Variable type`=="numeric" ~ "continuous",
                                       TRUE ~ `Variable type`),
           `Variable type` = 
             case_when(`Variable name` %in% binary_var~ "binary",
                                       TRUE ~ `Variable type`),
           n_unique = 
             case_when(`Variable name` %in% binary_var ~ as.integer(2),
                                TRUE ~ as.integer(n_unique)))
  TF="value_labels" %in% colnames(cb)
  if(TF==F){
    cb %<>% mutate(
      value_labels = case_when(`Variable name` %in% 
                                 binary_var~ "1, 2",  TRUE ~ NA_character_)) %>%
      as_tibble() %>%
      select(`Variable name`, `Variable label`= Label,
             `Variable type`,
             n_unique)
  } else {
    cb %<>% mutate(
      value_labels = case_when(`Variable name` %in% 
                                 binary_var~ "1, 2", TRUE ~ value_labels)) %>%
      as_tibble() %>%
      select(`Variable name`, `Variable label`= Label,
             `Variable type`,
             `Value labels` = value_labels,
             n_unique)
  }
}

is_categorical_var <- function(var){
  #(is.labelled(var) | is.factor(var) | is.character(var)) | (between(n_distinct(x[!is.na(x)]),3,20))
  (between(n_distinct(var[!is.na(var)]),3,20))
}
is_binary_var <- function(var){
  n_distinct(var[!is.na(var)])<=2
}
is_continous_var <- function(var){
  if (is.labelled(var)) {
    var = as.factor(var)
  }
  !(n_distinct(var[!is.na(var)])<=2 | is.factor(var))
}

datadic <- clean_data %>% datadic_var("")

write_xlsx(datadic, here("tables", "Assignment2_dd.xlsx"))
```

#Codebook
```{r}
library(epiDisplay)
library(readxl)
library(dplyr)
library(ggplot2)

codebook <- function(data, var, categorical = TRUE, print.values = TRUE) {
  # Ensure datadic contains the necessary columns
  if (!("Variable name" %in% colnames(datadic)) | !("Variable label" %in% colnames(datadic))) {
    stop("The required columns 'Variable name' or 'Variable label' are missing from datadic.")
  }

  # Pull the variable description
  var_description <- datadic %>% 
    filter(`Variable name` == var) %>% 
    pull(`Variable label`)

  # Check if 'Value labels' exists before pulling
  if ("Value labels" %in% colnames(datadic)) {
    var_values <- datadic %>% 
      filter(`Variable name` == var) %>% 
      pull(`Value labels`)
  } else {
    var_values <- NA
  }

  # Print out information to be included in the codebook
  print(str_c(var, ": ", var_description))
  if (!is.na(var_values) & print.values) {
    print(var_values)
  }

  if (categorical) {
    print(str_c("Frequency table of ", var))
    tab1(data[[var]], graph = FALSE, cum.percent = TRUE) %>% print()

    # Generate a bar chart
    print(str_c("Generating frequency chart for ", var))
    ggplot(data, aes(x = as.factor(.data[[var]]))) +
      geom_bar(fill = "grey", color = "black") +
      labs(title = paste("Frequency of", var), x = var, y = "Frequency") +
      theme_minimal() -> plot
    print(plot)

  } else {
    x <- data[[var]] %>% as.vector()
    na_count <- sum(is.na(x))
    na_perc <- round(na_count / length(x), digits = 3)
    print(str_c("NA Frequency: ", na_count, " (", 100 * na_perc, "%)"))

    # Generate a histogram for continuous variables
    print(str_c("Generating histogram for ", var))
    ggplot(data, aes(x = .data[[var]])) +
      geom_histogram(binwidth = 5, fill = "grey", color = "black") +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency") +
      theme_minimal() -> plot
    print(plot)
  }

  cat("\n\n")
}

codebook(clean_data, "age_cat", categorical = T)
codebook(clean_data, "gender", categorical = T)
codebook(clean_data, "race", categorical = T)
codebook(clean_data, "spd_lvl", categorical = T)
codebook(clean_data, "spd_month", categorical = T)
codebook(clean_data, "spd_year", categorical = T)
codebook(clean_data, "edu_lvl", categorical = T)
codebook(clean_data, "poverty_lvl", categorical = T)
codebook(clean_data, "marital_3cat", categorical = T)
codebook(clean_data, "marital_2cat", categorical = T)
codebook(clean_data, "subuse_mental", categorical = T)
codebook(clean_data, "kids", categorical = T)
codebook(clean_data, "binge", categorical = T)

```

#Table one
```{r}
listVars <- c("age_cat", "gender", "race","spd_lvl","spd_month","spd_year","edu_lvl","poverty_lvl", "subuse_mental", "kids","binge")
catVars <- c("age_cat", "gender", "race","spd_lvl","spd_month","spd_year","edu_lvl","poverty_lvl", "subuse_mental", "kids","binge")  
table1.overall <- CreateTableOne(vars = listVars, data = clean_data, factorVars = catVars,includeNA = TRUE)
a <- print(table1.overall, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
write.csv(a, "~/Downloads/table1.csv")

# stratified by marital status - 2 categories
table1.stratified <- CreateTableOne(vars = listVars, strata = "marital_2cat", data = clean_data, factorVars = catVars, includeNA = TRUE)
b <- print(table1.stratified, showAllLevels = TRUE, formatOptions = list(big.mark = ","))
write.csv(b, "~/Downloads/table1_stratified.csv")



```


#Table two: marital - 2 cat
## spd in the past month: crude
```{r}
clean_data$spd_month_recode <- ifelse(clean_data$spd_month == "Yes", 1, 0)
clean_data$spd_month_recode <- as.numeric(clean_data$spd_month_recode)
mod.crude <- glm(
  spd_month_recode ~ marital_2cat,
  data = clean_data,
  family = binomial(),
)
coef_summary <- summary(mod.crude)$coefficients
or <- exp(coef(mod.crude)) 
ci <- exp(confint(mod.crude))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}
```

## spd in the past month: adjusted
```{r}
mod.adj <- glm(
  spd_month_recode ~ marital_2cat + age_cat + gender + race + edu_lvl + poverty_lvl + subuse_mental,
  data = clean_data,
  family = binomial(),
)

coef_summary <- summary(mod.adj)$coefficients
or <- exp(coef(mod.adj)) 
ci <- exp(confint(mod.adj))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```


# spd in the past year - crude
```{r}
clean_data$spd_year_recode <- ifelse(clean_data$spd_year == "Yes", 1, 0)
clean_data$spd_year_recode <- as.numeric(clean_data$spd_year_recode)

mod.crude <- glm(
  spd_year_recode ~ marital_2cat,
  data = clean_data,
  family = binomial(),
)
coef_summary <- summary(mod.crude)$coefficients
or <- exp(coef(mod.crude)) 
ci <- exp(confint(mod.crude))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```


# spd in the past year - adjusted
```{r}
clean_data$spd_year_recode <- ifelse(clean_data$spd_year == "Yes", 1, 0)
clean_data$spd_year_recode <- as.numeric(clean_data$spd_year_recode)

mod.adj <- glm(
  spd_year_recode ~ marital_2cat + age_cat + gender + race + edu_lvl + poverty_lvl + subuse_mental,
  data = clean_data,
  family = binomial(),
)

coef_summary <- summary(mod.adj)$coefficients
or <- exp(coef(mod.adj)) 
ci <- exp(confint(mod.adj))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```
#Sensitivity: marital - 3 cat
## spd in the past month: crude
```{r}
clean_data$spd_month_recode <- ifelse(clean_data$spd_month == "Yes", 1, 0)
clean_data$spd_month_recode <- as.numeric(clean_data$spd_month_recode)

mod.crude <- glm(
  spd_month_recode ~ marital_3cat,
  data = clean_data,
  family = binomial(),
)
coef_summary <- summary(mod.crude)$coefficients
or <- exp(coef(mod.crude)) 
ci <- exp(confint(mod.crude))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}
```

## spd in the past month: adjusted
```{r}
mod.adj <- glm(
  spd_month_recode ~ marital_3cat + age_cat + gender + race + edu_lvl + poverty_lvl + subuse_mental,
  data = clean_data,
  family = binomial(),
)

coef_summary <- summary(mod.adj)$coefficients
or <- exp(coef(mod.adj)) 
ci <- exp(confint(mod.adj))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```

# spd in the past year - crude
```{r}
clean_data$spd_year_recode <- ifelse(clean_data$spd_year == "Yes", 1, 0)
clean_data$spd_year_recode <- as.numeric(clean_data$spd_year_recode)

mod.crude <- glm(
  spd_year_recode ~ marital_3cat,
  data = clean_data,
  family = binomial(),
)
coef_summary <- summary(mod.crude)$coefficients
or <- exp(coef(mod.crude)) 
ci <- exp(confint(mod.crude))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```
# spd in the past year - adjusted
```{r}
clean_data$spd_year_recode <- ifelse(clean_data$spd_year == "Yes", 1, 0)
clean_data$spd_year_recode <- as.numeric(clean_data$spd_year_recode)

mod.adj <- glm(
  spd_year_recode ~ marital_3cat + age_cat + gender + race + edu_lvl + poverty_lvl + subuse_mental,
  data = clean_data,
  family = binomial(),
)

coef_summary <- summary(mod.adj)$coefficients
or <- exp(coef(mod.adj)) 
ci <- exp(confint(mod.adj))  
for (i in seq_along(or)) {
  cat(names(or)[i], ": ", sprintf("%.2f (%.2f, %.2f)", or[i], ci[i, 1], ci[i, 2]), "\n")
}

```
# cases number
```{r}
tab1 <- table(clean_data$spd_month, clean_data$marital_2cat, 
             dnn = c("SPD last month", "Marital 2 Cat")) ;tab1

tab2 <- table(clean_data$spd_year, clean_data$marital_2cat, 
             dnn = c("SPD last year", "Marital 2 Cat")) ; tab2

tab3 <- table(clean_data$spd_month, clean_data$marital_3cat, 
             dnn = c("SPD last month", "Marital 3 Cat")) ;tab3

tab4 <- table(clean_data$spd_year, clean_data$marital_3cat, 
             dnn = c("SPD last year", "Marital 3 Cat")) ;tab4

```
